---
title: "Class_Project"
author: "Camille Berardone"
date: "2025-02-28"
output: html_document
editor_options: 
  chunk_output_type: console
---

#load all necessary libraries
```{r}
library(tidyr)  #for data reformatting
library(dplyr)  #for data reformatting
library(mobr)   #for data reformatting
library(openxlsx) #for creating excel files
library(vegan)   #for PCA analysis
library(ggplot2) #for PCA visualization
library(ggpubr)  #for PCA confidence ellipses
library(ggrepel)  #for PCA plot sample names
#install.packages("ggordiplots")
library(ggordiplots)  #for creating proper 95% CI ellipses on PCA
```

#SECTION 1: DATA REFORMATTING##################################################

#reformat raw data into long format
```{r}
dat_lkup <- read.csv('./data/Raw_Data_Protein_wSC.csv', header = FALSE,
                     nrows = 2)
dat_lkup <- t(dat_lkup[ , -(1:4)])
dat_lkup <- data.frame(dat_lkup)
names(dat_lkup) <- c('group', 'sample')
#head(dat_lkup)
dat_lkup$sample <- paste("X", dat_lkup$sample, sep='')

dat <- read.csv('./data/Raw_Data_Protein_wSC.csv', skip = 1)
dat[1:5, 1:5]

dat_lg <- pivot_longer(dat, cols = starts_with("X"), names_to = "sample")
dim(dat_lg)
#head(dat_lg)

dat_lg <- left_join(dat_lg, dat_lkup)
#head(dat_lg)
#View(dat_lg)

#write.csv(dat_lg, "transposed_raw_data.csv", row.names = FALSE)
```

#SECTION 2: RUNNING STATISTICAL ANALYSES#######################################

##MANN-WHITNEY TESTS
#OC2_0 vs OC4_0 all protein data - mann whitney u test
```{r}
#load in the data
all_mann <- read.csv("./data/wsc_normality.csv")
#View(all_mann)

#mann-whitney u test
pvals_test <- sapply(1:nrow(all_mann), function(i) {
  wilcox.test(as.numeric(all_mann[i, 5:13]), 
              as.numeric(all_mann[i, 23:37]), 
              paired = FALSE, alternative = "two.sided")$p.value
})

all_mann$p.value_OC2_0vsOC4_0 <- pvals_test
#View(all_mann)

#BH correction
padj_BH_OC2_0vsOC4_0 <- p.adjust(all_mann$p.value_OC2_0vsOC4_0, method = "BH")

#combine p values on sheet
all_mann<- cbind(all_mann, "padj_BH_OC2_0vsOC4_0"=padj_BH_OC2_0vsOC4_0)

#View the updated dataframe with Mann-Whitney U test results
#View(all_mann)
```

#OC2_1 vs OC4_1 all protein data - mann whitney u test
```{r}
#double check previous OC2_0 vs OC4_0 p-values are still there
View(all_mann)  #yes, they're still there

#mann-whitney u test
pvals_test <- sapply(1:nrow(all_mann), function(i) {
  wilcox.test(as.numeric(all_mann[i, 4:22]), 
              as.numeric(all_mann[i, 38:52]), 
              paired = FALSE, alternative = "two.sided")$p.value
})

all_mann$p.value_OC2_1vsOC4_1 <- pvals_test
#View(all_mann)

#BH correction
padj_BH_OC2_1vsOC4_1 <- p.adjust(all_mann$p.value_OC2_1vsOC4_1, method = "BH")

#combine p values on sheet
all_mann<- cbind(all_mann, "padj_BH_OC2_1vsOC4_1"=padj_BH_OC2_1vsOC4_1)

#View the updated dataframe with Mann-Whitney U test results
#View(all_mann)

#export data to excel .xlsx file
library(openxlsx)
#write.xlsx(all_mann, "04_02_mann_whitney_all_proteins_pandpadj_results.xlsx", rowNames = FALSE)
```

##PCA ANALYSIS
#run a PCA on the data
```{r}
#load the data
newpca <- read.csv("./data/new_PCA_data.csv")

#view the first few rows
#head(newpca)

#extract numeric data only
numeric_data <- newpca[, 3:ncol(newpca)]
#head(numeric_data)

#run the PCA
pca_result <- rda(sqrt(numeric_data), scale = FALSE) 
  #unsure if scaling protein abu is necessary
summary(pca_result)
#?rda

#extract PCA scores
pca_scores <- scores(pca_result, display = "sites")

#add back the group information
pca_df <- data.frame(pca_scores, Group = newpca$Group)

#visualize with 95% confidence ellipses and add sample names to the plot
#changing color scale to accommodate for colorblind friendly palettes
ggplot(pca_df, aes(x = PC1, y = PC2, color = Group, fill = Group)) +
  geom_point(size = 3, alpha = 0.7) +
  stat_ellipse(level = 0.95, geom = "polygon", alpha = 0.2) +
  geom_text_repel(aes(label = newpca$Sample), size = 4, max.overlaps = Inf) +
  #Add sample names
  scale_color_manual(values = c(
    "OC2_0" = "#D81B60",
    "OC2_1" = "#1E88E5",
    "OC4_0" = "#FFC107",
    "OC4_1" = "#004D40"
  )) +
  scale_fill_manual(values = c(
    "OC2_0" = "#D81B60",
    "OC2_1" = "#1E88E5",
    "OC4_0" = "#FFC107",
    "OC4_1" = "#004D40"
  )) +
  theme_classic() +
  labs(title = "PCA with 95% Confidence Intervals") +
  xlim(-5, 4) +
  ylim(-4, 5)
```

#use a permanova along with PCA for some statistical results
```{r}
#perform a permanova
permanova_result <- adonis2(numeric_data ~ Group, data = newpca,
                            method = "bray", permutations = 999)
permanova_result

#R^2 = 0.29851
#F = 6.2413
#p < 0.05
```

##RDA ANALYSIS
#run direct ordination with RDA - forces the x axis apart
```{r}
#run the rda
rda_result <- rda(sqrt(numeric_data) ~ newpca$Group)
summary(rda_result)

#look at adjusted R^2 value: 0.2669858; 27% variation explained by the groupings
RsquareAdj(rda_result)

anova(rda_result) #the model significantly explains the data
#F statistic = 6.7063

#create an rda plot
plot(rda_result, display = c('cn','sp','wa'))

#add group names to plot
text(rda_result, display = 'cn')
ordiellipse(rda_result, newpca$Group, display = 'sites')
```

#make the RDA plot look prettier
```{r}
#load the data
newpca <- read.csv("./data/new_PCA_data.csv")

#extract numeric data only
numeric_data <- newpca[, 3:ncol(newpca)]
#head(numeric_data)

# Run RDA
rda_result <- rda(sqrt(numeric_data) ~ newpca$Group)
summary(rda_result)

#relabel the labels within "Group" within the dataset
newpca$Group <- factor(newpca$Group, 
                       levels = c("OC2_0", "OC2_1", "OC4_0", "OC4_1"),
                       labels = c("OC2 Admission", "OC2 Rehabilitation",
                                  "OC4 Admission", "OC4 Rehabilitation"))

#extract site scores (samples)
sites <- scores(rda_result, display = "sites")
sites_df <- as.data.frame(sites)
sites_df$Group <- newpca$Group

#summarize to find group centers
group_centers <- sites_df %>%
  group_by(Group) %>%
  summarize(RDA1 = mean(RDA1), RDA2 = mean(RDA2))

#plot
ggplot(sites_df, aes(x = RDA1, y = RDA2, color = Group)) +
  geom_point(size = 3) +  # samples
  stat_ellipse(type = "norm", linetype = 1, level = 0.95) + 
    #ellipse around groups
  geom_text(data = group_centers, aes(label = Group), color = "black",
            fontface = "bold", size = 4) + # group labels
  theme_minimal() +
  labs(x = "RDA1",
       y = "RDA2") +
  theme(
    text = element_text(size = 14),
    plot.title = element_text(hjust = 0.5),
    legend.position = "right"
  ) +
  scale_color_manual(values = c(
    "OC2 Admission" = "#D81B60",
    "OC2 Rehabilitation" = "#1E88E5",
    "OC4 Admission" = "#FFC107",
    "OC4 Rehabilitation" = "#004D40"
  )) +  #colorblind colors
  geom_vline(xintercept = 0, linetype = "dotted", color = "black") +
      # dotted vertical line at x = 0
  geom_hline(yintercept = 0, linetype = "dotted", color = "black")
      # dotted horizontal line at y = 0
```

##BARPLOT
#create barplot for average number of proteins for each oiling category
```{r}
average <- read.csv("data/new_avg_ocs_proteins.csv")
average
average$count <- as.numeric(average$count)
str(average)


# Summarize data to calculate mean and standard deviation
summary_data <- average %>%
  group_by(OC) %>%
  summarize(
    mean_count = mean(count),
    sd_count = sd(count)
  )

#plot with bars and error bars and change the colors of the bars (colorblind)
ggplot(summary_data, aes(x = OC, y = mean_count, fill = OC)) + 
  geom_bar(stat = "identity", color = "black", width = 0.7) + 
  geom_errorbar(aes(ymin = mean_count - sd_count, ymax = mean_count + sd_count), 
                width = 0.2) + 
  labs(x = "Oiling Category", y = "Mean Protein Count (±SD)") + 
  scale_fill_manual(values = c(
    "OC2_Admission" = "#D81B60",
    "OC2_Rehabilitation" = "#1E88E5",
    "OC4_Admission" = "#FFC107",
    "OC4_Rehabilitation" = "#004D40"
  )) +
  theme_classic() + 
  theme(legend.position = "none") +
  theme(axis.title.x = element_text(size = 20, face = "bold"),
        axis.title.y = element_text(size = 20, face = "bold"),
        axis.text.x = element_text(size = 12,face = "bold"),
        axis.text.y = element_text(size = 12,face = "bold")) + 
  scale_y_continuous(expand = c(0, 0), limits = c(0, max(average$count) * 1.05),
                     breaks = seq(0, max(average$count), by = 50))
```

#anova with tukey posthoc for average number of proteins per group
```{r}
#read in the data
avgaov <- read.csv("data/new_avg_ocs_proteins.csv")
avgaov

anova <- aov(count~OC, data=avgaov)
summary(anova)

tukey <- TukeyHSD(anova)
tukey

#OC2_0 vs OC2_1 = significant
#OC2_0 vs OC4_0 = significant
#OC2_1 vs OC4_1 = not significant
#OC4_0 vs OC4_1 = significant
```

#SECTION 3: FUTURE ANALYSES####################################################

#mobr diversity comparison - will look into after project completion
```{r}
pro_mob_in <- make_mob_in(dat[ , 5:ncol(dat)], plot_attr = dat[ , 1:4])
pro_mob_in

plot_rarefaction(pro_mob_in, group_var = )
```

#reanalyzing 95% confidence interval ellipses on PCA - will look into after
project completion
```{r}
#note: using code comments from abbey's review file
#create new pca plot with 95% confidence intervals
pca_plot <- gg_ordiplot(pca_result, groups = pca_df$Group, kind = "se",
                        conf = 0.95, pt.size = 3)

# Add sample names into the site scores
pca_plot$data$sites$SampleName <- newpca$Sample

# Now plot
pca_plot$plot +
  geom_text(data = pca_plot$data$sites, 
            aes(x = x, y = y, label = SampleName),
            #now label is inside the data
            vjust = -1, size = 3) +
  ggtitle("PCA of Samples") +
  theme_classic()

# 1. Turn pca_plot$data$sites into a real data frame
site_scores <- as.data.frame(pca_plot$data$sites)

# 2. Add your sample names to the new data frame
site_scores$SampleName <- newpca$Sample

# 3. Now plot using site_scores
pca_plot$plot +
  geom_text(data = site_scores, 
            aes(x = PC1, y = PC2, label = SampleName), 
            vjust = -1, size = 3) +
  ggtitle("PCA of Samples") +
  theme_classic()
```

#identify key proteins in the data
    #notes to self on how to do this...
      #could use datasheet with calculated empai molar ratios and separate from
      smallest to largest
      #organize by top 10 in each
      #look for similar and different proteins. what is the overall function of
      these proteins?
```{r}
#note2: use code from assignment that organized species from most to
#least abundant
```

#run direct ord with CCA - look at proportional (rather than absolute changes)
```{r}
#run the cca
cca_result <- cca(sqrt(numeric_data) ~ newpca$Group)

#look at adjusted R^2 value: 0.1449003; 14% variation explained by the groupings
RsquareAdj(cca_result)

anova(cca_result) #the model significantly explains the data

#create a cca plot
plot(cca_result, display = c('cn','sp'))

#add group names to plot - not the best display of the data
text(cca_result, display = 'cn')
```
